HMMotion: Using tracking data to predict coverage strategies in American Football

### 2024-12-20 submission.Rmd
.

Choosing between man- and zone coverage is one of the most important strategic decisions a defensive coordinator has to make prior to each offensive play in American football. While opposing offensive coordinators and quarterbacks try to identify these defensive strategies visually, the rise of tracking data provides another opportunity to infer the hidden defensive tactics.

Previous approaches attempted to predict zone- or man coverage. One example can be seen in the below screenshot which displays a snapshot from the Amazon Prime broadcast of the week 12 match between the Pittsburgh Steelers and Cleveland Browns of the current NFL season, including a prediction by Amazon's NFL Next Gen Stats model whether man- or zone coverage will be played. Although the full model and its details are not public, the live broadcast prediction mostly focuses on specific plays without any pre-snap motion. We proceeed similarly, however, in this project, we do not stop there. In contrast, we also depict player movements before the snap, i.e.\ pre-snap motion, and exploit this additional information using hidden Markov models (HMMs). By modeling the trajectories of defenders depending on hidden states, which represent the offensive players, we infer probabilities for each offensive player being guarded by the specific defenders in every instance of a play that contains player movements. We map this high-dimensional time series information to a single number for each play by calculating the entropy, thus, ending up with a measure of the degree of uncertainty. Incorporating this information into the existing pre-motion model, we can show that the AUC and the detection accuracy are increased. In this way, we provide a data-driven framework that is able to assess the efficiency of pre-snap motion to reveal the defensive strategy, enabling real-time tactical insights for coaches.

>>>>

Predicting whether a defense is playing man or zone coverage has long been a focal point in football analytics. A notable example is Amazon's NFL Next Gen Stats model, which delivers predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns. These predictions, however, are often limited to analyzing specific plays without incorporating pre-snap motion which is a crucial element of modern offensive strategies.

Our project takes this analysis a step further. While we similarly predict man or zone coverage when the teams are set, we further leverage the additional information of pre-snap player movements. Using hidden Markov models (HMMs), we model defenders' trajectories based on hidden states, representing the offensive players they may be guarding. This probabilistic approach allows us to infer dynamic coverage patterns during plays with pre-snap motion.

To exploit this complex, high-dimensional data into actionable insights, we calculate the entropy of the state probabilities, producing a single measure of uncertainty for each play. Incorporating these as covariates into the existing pre-motion model significantly improves both the AUC and detection accuracy. In doing so, we present a data-driven framework that evaluates the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.
